Miniproject - Combinatorics

### 2021-12-15 Combinatorics/submission.tex
Added final thoughts to combinatorics

\begin{proposition}
    \label{prop:core_of_multisets}
    Assume that $\mathcal{C}_{k, d}$ satisfies a ``scaled variant'' of condition (3) of Prop.~\ref{prop:characterization_lym}, namely:
    There exists a (finite) sequence $C'_1, ..., C'_t$ of maximal chains such that for a chain $C'$ chosen uniformly at random among them, we have
    \begin{equation*}
        \Pr[ (n_0, ..., n_k) \in C' ] = \frac {|P(n_0, ..., n_k)|} {|L_{\sum_i i n_i}|}
    \end{equation*}
    Then $\mathcal{P}_{k, d}$ satisfies the conditions of Prop.~\ref{prop:characterization_lym}.
\end{proposition}
\begin{proof}
    Denote the layers of $\mathcal{P}_{k, d}$ by $L_k$ and the layers of $\mathcal{C}_{k, d}$ by $L'_k$.

@@

Hence, we get the following statement.
\begin{proposition}
    Let $P$ be a graded poset. Define the \emph{core} $\mathcal{C}(P)$ of $P$ as the set of orbits under the action of $\mathrm{Aut}(P)$ 
    \begin{equation*}
        \mathcal{C}(P) := \{ \mathrm{Aut}(P).x \ | \ x \in P \}
    \end{equation*}
    with the induced partial ordering from the elements of each orbit, i.e.
    \begin{equation*}
        A < B \ :\Leftrightarrow \ \exists a \in A, b \in B: a < b
    \end{equation*}
    for $A, B \in \mathcal{C}(P)$.

    If there exists a sequence of maximal chains $C'_1, ..., C'_t$ in $\mathcal{C}(P)$ such that for a chain $C$ chosen uniformly at random from among them, have
    \begin{equation*}
        \Pr[A \in C] = \frac {|A|} {|L_{r(A)}|}
    \end{equation*}
    for all $A \in \mathcal{C}(P)$, then $P$ satisfies the conditions of Prop.~\ref{prop:characterization_lym}.
\end{proposition}
\begin{proof}
    Exactly as above.
\end{proof}
I still do not know how to use this to prove that $\mathcal{P}_{k, d}$ does indeed satisfy Prop.~\ref{prop:characterization_lym}, but at least it provides a way for an even faster algorithm to test it for specific $k, d$.
The idea is just that we can adjust the LP algorithm from Lemma~\ref{prop:lp_algorithm} to check the scaled variant, and so we get a polynomial time algorithm in $|\mathcal{C}_{k, d}|$.
In particular, we have the following.
\begin{lemma}
    Let $L_{l - 1} = \{u_1, ..., u_m\}$ and $L_l = \{v_1, ..., v_n\}$ be two consecutive layers of $\mathcal{C}_{k, d}$.
    Assume there exists a matrix $A = (a_{ij}) \in \Q_{\geq 0}^{m \times n}$ such that 
    \begin{equation*}
        1^T A 1 = 1
    \end{equation*}
    and for all $i, j$
    \begin{equation*}
        e_i^T A 1 = \frac {|P(u_i)|} {\sum_{u \in L_{l - 1}} |P(u)|} \quad \text{and} \quad 1^T A e_j = \frac {|P(v_j)|} {\sum_{v \in L_l} |P(v)|}
    \end{equation*}
    and $a_{ij} = 0$ whenever $u_i \not< v_j$ holds.

@@

After quite a long time trying to prove that $\mathcal{P}_{k, d}$ satisfies an LYM-inequality analogue, I finally gave up and searched the literature.
There I found that $\mathcal{P}_{k, d}$ is also called divisor lattice and indeed we have the following statement.
\begin{proposition}
    $\mathcal{P}_{k, d}$ satisfies the conditions of Prop.~\ref{prop:characterization_lym}, i.e. an LYM-inequality.
\end{proposition}
\begin{proof}
    See e.g. \cite[Theorem 4.2.3]{anderson}.
\end{proof}
Interestingly, I have also found a variant of the LYM inequality for the poset $\Pi_m$ that we studied earlier.
\begin{proposition}
    Let $\mathcal{A} \subseteq \Pi_m$ be a nonempty antichain. Then define
    \begin{equation*}
        a_{n_1, ..., n_k} := |\bigl\{ A \in \mathcal{A} \ \bigm| \ |A| = k, \ |\{ S \in A \ | \ |S| = j \}| = n_j \bigr\}|
    \end{equation*}
    the number of partitions $A \in \mathcal{A}$ such that the classes in $A$ have sizes $n_1, ..., n_k$.
    Then
    \begin{equation*}
        \sum_{k \geq 1} \sum_{n_1, ..., n_k} \frac {a_{n_1, ..., n_k} k!} {{m \choose n_1, ..., n_k} {m - 1 \choose k - 1}} \leq 1
    \end{equation*}
\end{proposition}
\begin{proof}
    See \cite{lym_partitions}.
\end{proof}

>>>>

\begin{proposition}
    \label{prop:core_of_multisets}
    Assume that $\mathcal{C}_{k, d}$ satisfies a ``scaled variant'' of condition (3) of Prop.~\ref{prop:characterization_lym}, namely:
    There exists a (finite) sequence $C'_1, ..., C'_t$ of maximal chains such that for a chain $C'$ chosen uniformly at random among them, we have
    \begin{equation*}
        \Pr[ (n_0, ..., n_k) \in C' ] = \frac {|P(n_0, ..., n_k)|} {|L_{r(n_0, ..., n_k)}|}
    \end{equation*}
    Then $\mathcal{P}_{k, d}$ satisfies the conditions of Prop.~\ref{prop:characterization_lym}.
\end{proposition}
\begin{proof}
    Denote the layers of $\mathcal{P}_{k, d}$ by $L_k$ and the layers of $\mathcal{C}_{k, d}$ by $L'_k$.

@@

Hence, we get the following statement.
\begin{proposition}
    Let $P$ be a graded poset. Define the \emph{core} $\mathcal{C}(P)$ of $P$ as the set of orbits under the action of $\mathrm{Aut}(P)$ 
    \begin{equation*}
        \mathcal{C}(P) := \{ \mathrm{Aut}(P).x \ | \ x \in P \}
    \end{equation*}
    with the induced partial ordering from the elements of each orbit, i.e.
    \begin{equation*}
        A < B \ :\Leftrightarrow \ \exists a \in A, b \in B: a < b
    \end{equation*}
    for $A, B \in \mathcal{C}(P)$.
    Note that then $\mathcal{C}(P)$ is a graded poset.

    If there exists a sequence of maximal chains $C'_1, ..., C'_t$ in $\mathcal{C}(P)$ such that for a chain $C$ chosen uniformly at random from among them, have
    \begin{equation*}
        \Pr[A \in C] = \frac {|A|} {|L_{r(A)}|}
    \end{equation*}
    for all $A \in \mathcal{C}(P)$, then $P$ satisfies the conditions of Prop.~\ref{prop:characterization_lym}.
\end{proposition}
\begin{proof}
    Exactly as in Prop.~\ref{prop:core_of_multisets}.
\end{proof}
I still do not know how to use this to prove that $\mathcal{P}_{k, d}$ does indeed satisfy Prop.~\ref{prop:characterization_lym}, but at least it provides a way for an even faster algorithm to test it for specific $k, d$.
The idea is just that we can adjust the LP algorithm from Lemma~\ref{prop:lp_algorithm} to check the scaled variant, and so we get a polynomial time algorithm in $|\mathcal{C}_{k, d}|$.
In particular, we have the following.
\begin{lemma}
    Let $L_{l - 1} = \{u_1, ..., u_m\}$ and $L_l = \{v_1, ..., v_n\}$ be two consecutive layers of $\mathcal{C}_{k, d}$.
    Assume there exists a matrix $A = (a_{ij}) \in \Q_{\geq 0}^{m \times n}$ such that 
    \begin{equation*}
        1^T A 1 = 1
    \end{equation*}
    and for all $i, j$
    \begin{equation*}
        e_i^T A 1 = \frac {|P(u_i)|} {\sum_{u \in L_{l - 1}} |P(u)|} \quad \text{and} \quad 1^T A e_j = \frac {|P(v_j)|} {\sum_{v \in L_l} |P(v)|}
    \end{equation*}
    and $a_{ij} = 0$ whenever $u_i \not< v_j$ holds.

@@

After quite a long time trying to prove that this holds generally, I finally gave up and searched the literature.
There I found that $\mathcal{P}_{k, d}$ is also called divisor lattice and indeed we have the following statement.
\begin{proposition}
    $\mathcal{P}_{k, d}$ satisfies the conditions of Prop.~\ref{prop:characterization_lym}, i.e. an LYM-inequality.
\end{proposition}
\begin{proof}
    See e.g. \cite[Theorem 4.2.3]{anderson}.
\end{proof}
Interestingly, I have also found a variant of the LYM inequality (``pseudo-LYM inequality'') for the poset $\Pi_m$ that we studied earlier.
\begin{proposition}
    Let $\mathcal{A} \subseteq \Pi_m$ be a nonempty antichain. Then define
    \begin{equation*}
        a(n_1, ..., n_k) := |\bigl\{ A \in \mathcal{A} \ \bigm| \ |A| = k, \ |\{ S \in A \ | \ |S| = j \}| = n_j \bigr\}|
    \end{equation*}
    the number of partitions $A \in \mathcal{A}$ such that the classes in $A$ have sizes $n_1, ..., n_k$.
    Then
    \begin{equation*}
        \sum_{k \geq 1} \sum_{n_1, ..., n_k} \frac {a(n_1, ..., n_k) k!} {{m \choose n_1, ..., n_k} {m - 1 \choose k - 1}} \leq 1
    \end{equation*}
\end{proposition}
\begin{proof}
    See \cite{lym_partitions}.
\end{proof}
Note that the total number of partitions $\pi$ such that the classes in $\pi$ have sizes $n_1, ..., n_k$ is given by an expression involving the multinomial coefficient, namely
\begin{equation*}
    |\bigl\{ \pi \in \Pi_m \ \bigm| \ |\pi| = k, |\{ S \in \pi \ | \ |S| = j\}| = n_j \bigr\}| = {m \choose n_1, ..., n_k} \underbrace{ \prod_i \frac 1 {|\{ j \ | \ n_j = i\}|!} }_{=: \gamma(n_1, ..., n_k)}
\end{equation*}
Intuitively, the reason is that there are $m!$ ways to order the elements $[m]$, and after splitting such a sequence in chunks of sizes $n_1, ..., n_k$, there are $n_1! \cdot ... \cdot n_k!$ ways to order the element within the chunks.
Hence the multinomial coefficient
\begin{equation*}
    {m \choose n_1, ..., n_k}
\end{equation*}
gives the number of ways to distribute the elements of $[m]$ into ordered buckets of sizes $n_1, ..., n_k$.
A partition corresponds to unordered buckets, i.e. swapping chunks of equal size does not change the partition.
There are $1/\gamma(n_1, ..., n_k)$ ways to do so, and hence we get the expression mentioned above.

Furthermore, the number $a(n_1, ..., n_k)$ and the underlying set do not depend on the order of $n_1, ..., n_k$.
Hence, the partitions $A \in \mathcal{A}$ such that the classes of $A$ have sizes $n_1, ..., n_k$ might be counted multiple times.
Note that the number of different orderings of $n_1, ..., n_k$ is given by $k!/\gamma(n_1, ..., n_k)$.
Hence
\begin{equation*}
    \sum_{n_1, ..., n_k} \frac 1 {k!} {m \choose n_1, ..., n_k} = |L_{m - k}|
\end{equation*}
just counts all the partitions in one layer of $\Pi_m$.

Now note also that
\begin{equation*}
    {m - 1 \choose k - 1}
\end{equation*}
is exactly the number of different sequences $n_1 \leq ... \leq n_k$ by the ``stars and bars'' method.
Then one can see that the expression
\begin{equation*}    
    \sum_{k \geq 1} \sum_{n_1, ..., n_k} \frac {a(n_1, ..., n_k) k!} {{m \choose n_1, ..., n_k} {m - 1 \choose k - 1}} = \sum_{k \geq 1} \frac 1 {{m - 1 \choose k - 1}} \sum_{n_1, ..., n_k} \frac {a(n_1, ..., n_k)} {\frac 1 {k!} {m \choose n_1, ..., n_k}}
\end{equation*}
is very similar to the LYM inequality.
The only difference is that elements elements corresponding to different class sizes $n_1, ..., n_k$ might receive slightly different weight, depending on how many duplicate $n_i$ occur in that sequence.

Finally, there is still one interesting thought I had while thinking about the above inequality.
Note that the decomposition
\begin{equation*}
    \Pi_m = \bigcup_{k \geq 1, \ n_1 \leq ... \leq n_k} \{ \pi \in \Pi_m \ | \ |\pi| = k, |\{ S \in \pi \ | \ |S| = j \}| = n_j \}
\end{equation*}
corresponds exactly to the core $\mathcal{C}(\Pi_m)$ of $\Pi_m$ as discussed above.
In particular, the decomposition is exactly
\begin{equation*}
    \Pi_m = \bigcup \ \{ \mathrm{Aut}(\Pi_m).\pi \ | \ \pi \in \Pi_m \}
\end{equation*}
